---
title: "Project1-ChessDataAnalysis"
author: "Ross Boehme"
date: "2023-02-18"
output: html_document
---

## **DATA 607 Project 1**

Take chess tournament results text file, and create an R Markdown document which generates a structured CSV file. Output CSV should follow the format: 
1. Player’s Name
2. Player’s State
3. Total Number of Points
4. Player’s Pre-Rating
5. Average Pre Chess Rating of Opponents

For example, the first player would be
Gary Hua, ON, 6.0, 1794, 1605


#### Loading Packages

```{r}
library(tidyverse)
```

#### Text Import

Because I can't import the text from Blackboard (due to it being behind a login wall), I uploaded the text file to my own Github, where I'll read it from.

Reading the file is somewhat difficult because it has an unconventional structure. The simplest solution I could find was using readLines. 

```{r}
chess <- readLines("https://raw.githubusercontent.com/rossboehme/DATA607/main/project1/data607-project1-chess.txt")
```


### **Scraping File With Regex**

My approach is to first remove unnecessary info and split the df into three groups based on the information a given line provided. This should make my regex easier. 

#### Removing Unnecessary Info; Preparing Data for Scraping

```{r}
#Remove first four rows
chess <- chess[-c(0:4)]
```

```{r}
#Splitting dataframe into three groups based on rows

#Rows 1, 4, 7, etc. contain Pair Number, Player Name, Total Points, Opponent info, 
#Aliasing "PairNamePointsOpp"
PairNamePointsOpp <- chess[seq(1, length(chess),3)]

#Rows 2, 5, 8, etc. contain Player State, Pre Rating. 
#Aliasing "StateRating"
StateRating <- chess[seq(2, length(chess),3)]

#Rows 3, 6, 9, etc. can be removed. Don't need to be saved.
```

#### Scraping Pair Number, Player's Name, Total Points, and Opponent Info
```{r}
#Note: I put my regexes on lines by themselves so I can visualize them easier

pair_num <- as.integer(unlist(str_extract_all(PairNamePointsOpp,
                                              "(?<=\\s{3,4})\\d{1,2}(?=\\s)"
                                              )))

name <- unlist(trimws(str_extract(PairNamePointsOpp,
                           "([[A-Z]]+\\s){2,3}"
                           )))

points <- as.numeric(unlist(str_extract(PairNamePointsOpp,
                             "\\d+\\.\\d+"
                             )))

opponents <- as.integer(unlist(str_extract_all(str_extract_all(PairNamePointsOpp,
                                                               "[[0-9]]+\\|"),"[[0-9]]+"
                                               )))
```
#### Scraping Player's State, Pre Rating
```{r}
state <- str_extract(StateRating, 
                     "[[A-Z]]+"
                     )

# I'll pull two versions of the "pre ratings": 
# 1) The full version includes the alphabetical characters e.g. "EZEKIEL HOUGHTON": 1641P17. I'll call this simply "pre-rating" 
# 2) However, since I need to average players opponents' Pre Ratings for my final product, I'll pull another adjusted ("adj") version which is only the numeric characters, and use that for my calculations
pre_rating <- unlist(trimws(str_extract(StateRating, 
                                                 "(?<=>)(\\s)?[0-9A-Z]{3,7}"
                                     )))

pre_rating_adj <- as.integer(unlist(trimws(str_extract(StateRating, 
                                                 "(?<=>)(\\s)?[0-9]{3,4}"
                                     ))))

```

#### Putting Scraped Information into Dataframe

```{r}
chess_cleaned <- data.frame(name,state,points,pre_rating)
```


### **Calculating Pre Chess Rating of Opponents; Adding to DF**

As a final step, I need to add the Average Pre Chess Rating of Opponents to my chess_cleaned df created above. There are 6.375 opponents for every player ("name"), therefore there are some players who played fewer than 7 matches. 

```{r}
#Average of 6.375 opponents for every player
length(opponents) / length(name)
```
As a solution, I'll create a matrix containing the pair numbers each player ("name") played, this means there will be NA values accounting for missed matches.

```{r}
col_df <- str_split(PairNamePointsOpp, pattern = "\\|",simplify = TRUE)

opp_matrix <- matrix(as.numeric(str_extract_all(col_df[,4:10], pattern = "..$")), ncol = 7)
```

I'll run a for loop over that matrix, averaging for each row (while skipping over NA values) the opponents' pre ratings.

```{r}
avg_opp_pre_rating <- c()

for(i in 1:nrow(opp_matrix)){
  avg_opp_pre_rating[i] <- round(mean(pre_rating_adj[opp_matrix[i,]], na.rm = TRUE),0)
}
```

Finally I'll add this column to my df.
```{r}
chess_cleaned$avg_opp_pre_rating = avg_opp_pre_rating
```


### **Showing Dataframe; Writing to CSV**
```{r}
chess_cleaned

#User should write to whatever path they want. For me, it's my desktop.
write.csv(chess_cleaned,"C:\\Users\\rossboehme\\Desktop\\chess.csv",row.names=FALSE)
```