---
  title: "Extra Credit - Windows Functions"
author: "Ross Boehme"
date: "2023-02-13"
output: html_document
---
  
<!-- Windows Functions Extra Credit -->
  
<!-- 1. Find a dataset that includes time series for two or more separate items.  For example, you could use end of day stock or cryptocurrency prices since Jan 1, 2022 for several instruments. -->
  
Reading in Citi Bike data. These data contain every Citi Bike trip taken in January 2023. 

```{r}
url <- "https://s3.amazonaws.com/tripdata/JC-202301-citibike-tripdata.csv.zip"
download.file(url,"202301-citibike-tripdata.csv.zip")
unzip("202301-citibike-tripdata.csv.zip")
bikes <- read.csv("202301-citibike-tripdata.csv")
```
```{r}
ncol(bikes)
column_names <- names(bikes)
```
These data currently includes 13 columns:
1) "ride_id" - unique identifier for each ride
2) "rideable_type" - values of "docked", "elecric", or "classic". docked bikes = trips where the bike didn't move (accidental unlocks). 
3) "started_at" - date and time of the start of the trip (d/m/yy hr:min)
4) "ended_at" - date and time of the end of the trip (d/m/yy hr:min)
5) "start_station_name" - Street name/s of location of start station
6) "start_station_id" - ID of the station where the bike was taken
7) "end_station_name" - Street name/s of location of end station
8) "end_station_id" - ID of the station where the bike was returned
9) "start_lat" - latitude of start station
10) "start_lng" - longitude of start station
11) "end_lat" - latitude of start station
12) "end_lng" - longitude of start station
13) "member_casual" - annual membership citi bike subscriber ("member") or anyone else "casual"

I don't need all these columns. I will remove everything but columns I will use in my analysis: rideable_type (bike types), started_at and ended_at (my time series data), as well as member_casual (rider type).

```{r}
bikes <- bikes[,c('rideable_type','member_casual','started_at','ended_at')]
```

Also I'll remove rows with a 'rideable_type' value of docked_bike since those are accidental unlocks where trip times aren't representative of actual trips.

```{r}
bikes <- bikes[bikes$rideable_type != 'docked_bike',]
```

Finally, before I perform my analysis, I'll load packages which I'll use for calculating my rolling averages and data manipulation. In addition, I'll and add a column "duration" which calculates the difference between the start_time and end_time for each trip.

```{r}
library(dplyr) #used to execute windows functions
library(zoo) #used to execute windows functions
library(lubridate) #used to format trip duration below
library(hms) #used to calculate trip duration below
library(purrr) #used to calculate rolling YTD average
```
```{r}
bikes <- mutate(bikes , trip_duration=as_hms(difftime(ended_at,started_at)))
```

<!-- 2. Use window functions (in SQL or dplyr) to calculate the year-to-date average and the six-day moving averages for each item. -->
  
In order to calculate daily moving averages, I need to group my data by trip start time. First I create my start_date column.

```{r}
bikes$start_date <- as.Date(bikes$started_at)
```

Currently, my data is organized with each observation (row) being an individual bike trip. To calculate my rolling averages, I'll instead aggregate my numeric series "trip_duration" by my 3 qualitative series rideable_type, member_casual, and my timeseries data start_date.

```{r}
bikes <- bikes %>% 
  group_by(start_date,rideable_type,member_casual) %>%
  summarise(trip_duration=sum(trip_duration))
```
For my YTD rolling average, I need to have a Year column, and the trip_duration column needs to temporarily be a data type of numeric. 
    
```{r}
bikes$start_year <- year(bikes$start_date)
bikes$trip_duration <- as.numeric(bikes$trip_duration)
```

Now I'll calculate YTD rolling average
```{r}
cumavg <- function(x) cumsum(x) / seq_along(x)
bikes <- transform(bikes, duration_ytdavg = ave(trip_duration, start_year, FUN = cumavg))
```

Then calculate 3 and 6 day rolling averages using rollmean from zoo. 
```{r}
bikes <- bikes %>%
    group_by(rideable_type,member_casual) %>% 
    mutate(duration_3dayavg = rollmean(trip_duration, k = 3, fill = NA),
            duration_6dayavg = rollmean(trip_duration, k = 6, fill = NA)) %>% 
    ungroup()
```

Changing to period data types for readability
```{r}
bikes <- bikes %>%
  mutate(trip_duration = seconds_to_period(trip_duration)) %>%
  mutate(duration_3dayavg = seconds_to_period(duration_3dayavg)) %>%
  mutate(duration_6dayavg = seconds_to_period(duration_6dayavg)) %>%
  mutate(duration_ytdavg = seconds_to_period(duration_ytdavg))
```

```{r}
write.csv(bikes,"C:\\Users\\rossboehme\\Desktop\\DATA607\\bikes-rolled.csv",row.names=FALSE)
```

3. Present your code in a three to five minute presentation  (or you may make a recording using screen-castomatic or another tool).

